---
title: Weekly Deaths in Northern Ireland
---

```{r setup, warning = FALSE}
# -------------------------
# Load required packages
# -------------------------
library(here)
library(dplyr)
library(plotly)
library(leaflet)
library(sf)
library(ggplot2)
library(RColorBrewer)
library(stringr)
library(ggrepel)
library(base64enc)
library(lubridate)

source(here("dashboard_data_prep.R"))
```


<!-- --------------------------------------
     PAGE HEADER: Publication date + actions
     Displays the latest week ending date and navigation buttons
--------------------------------------- -->

<div class="page-date-heading">
  <!-- Dynamic publication date pulled from R -->
  Week ending Friday, `r week_ending`
  
  <div class="publication-actions">
  
  <!-- Button: Navigate to About / methodology page -->
<div class="publication-actions">
  <button class="publication-action-button" type="button" onclick="window.location.href='about.html'">About</button>
  <button class="publication-action-button" type="button" onclick="sharePage()">Share</button>
  <button class="publication-action-button" type="button" onclick="window.location.href='weekly-deaths.pdf'">Download</button>
</div>
</div>
</div>
  
<details open class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">


<!-- Section title shown to users -->
<h4 class="accordion-title">
Provisional number of weekly deaths registered
</h4>


<!-- Accordion expand/collapse icon -->
<span class="accordion-symbol" aria-hidden="true"></span>
</div>
</summary>

<div class="card-body">


<!-- Grid layout for responsive cards -->
::: {.grid}


<!-- --------------------------------------
     CARD 1: Registered deaths
--------------------------------------- -->
::: {.g-col-12 .g-col-md-6}
::: {.card-padding .d-flex .h-100}
::: {.card .w-100}

::: {.card-header}
::: {style="text-align: left;"}


<!-- Card title -->
##### Deaths Registered
:::
:::

::: {.card-body .text-center}


<!-- Main metric: total deaths for latest week -->
<div class="fs-1 fw-bold" style="color:#00205b; text-align:left;">
`r reg_wdths`
</div>


<!-- Weekly change compared with previous week -->
<div class="fs-6 fw-bold" style="color:#000000; text-align:left;">
`r abs(reg_wdths_diff)` `r reg_wdths_arrow` deaths than previous week
</div>


<!-- Reference value from previous week -->
<div class="fs-6 text-muted" style="text-align:left;">
Previous week: `r reg_wdths_prev`
</div>

:::
:::
:::
:::


<!-- --------------------------------------
     CARD 2: Comparison with expected deaths
--------------------------------------- -->
::: {.g-col-12 .g-col-md-6}
::: {.card-padding .d-flex .h-100}
::: {.card .w-100}

::: {.card-header}
::: {style="text-align: left;"}


<!-- Card title -->
##### Comparison with expected deaths
:::
:::

::: {.card-body .text-center}


<!-- Absolute difference between registered and expected deaths -->
<div class="fs-1 fw-bold" style="color:#00205b; text-align:left;">
`r abs(diff_wdths)`
</div>


<!-- Indicator showing whether deaths are above/below expected -->
<div class="fs-6 fw-bold" style="color:#000000; text-align:left;">
`r diff_arrow` compared with expected weekly deaths
</div>


<!-- Expected deaths benchmark -->
<div class="fs-6 text-muted" style="text-align:left;">
Expected deaths: `r expect_wdths`
</div>

:::
:::
:::
:::

:::


</div>
</details>


<!-- --------------------------------------
     SECTION: 52-week trend (Registered vs Expected deaths)
     Displays weekly registered deaths (bars) and expected deaths (line)
     for the most recent 52 weeks
--------------------------------------- -->

<details open class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">

<!-- Section title shown to users -->
<h4 class="accordion-title">
Provisional number of weekly registered deaths and expected weekly deaths for the past 52 weeks
</h4>


<!-- Accordion expand/collapse icon -->
<span class="accordion-symbol" aria-hidden="true">
</span>
</div>
</summary>

<div class="card-body">


::: {.card-padding}
::: {.card}

<!-- Prompt mobile users to rotate screen for better viewing -->

<div class="rotate-screen">
  <p>Please rotate your screen for optimal experience</p>
</div>

::: {.card-body}

```{r fig_overall_expected_repeat, fig-format: svg, echo=FALSE, out.width="100%", fig.alt="Bar chart showing registered deaths by week  alongside a line for expected deaths. Registered deaths fluctuate over time and generally track close to expected deaths, with some variation above and below the expected values."}


# -------------------------
# Prepare data
# -------------------------

df_observe_excess <- df_observe_excess %>%
  dplyr::mutate(
    
    # Ensure week_date is in Date format
    week_date = as.Date(week_date)
  ) %>%
  dplyr::arrange(week_date) %>%
  
  # Filter to most recent 52 weeks
  dplyr::filter(
    week_date >= max(week_date, na.rm = TRUE) - lubridate::weeks(52)
  ) %>%
  dplyr::mutate(
    
    # Create sequential index for x-axis
    week_num = dplyr::row_number(),
    
    
    # Create offsets to align expected deaths line with bar edges (step-style line)
    bar_left = week_num - 0.50,
    bar_right = week_num + 0.40,
    
    # Extract month and year for axis labelling
    month = lubridate::floor_date(week_date, "month"),
    year = lubridate::year(week_date),
    month_label = format(week_date, "%b")
  )

# -------------------------
# Create custom x-axis ticks (monthly labels)
# -------------------------

month_ticks <- df_observe_excess %>%
  dplyr::group_by(month) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    
    # Identify where a new year starts
    year_marker = year != dplyr::lag(year, default = dplyr::first(year) - 1),
    
    # Add year to label when it changes
    x_tick_label = dplyr::if_else(
      year_marker,
      paste0(month_label, "<br>", year),
      month_label
    )
  )

# -------------------------
# Prepare expected deaths line (step-style)
# -------------------------
df_expected_outline <- df_observe_excess %>%
  dplyr::select(week_label, week_date, EXPDTHS, bar_left, bar_right) %>%
  
  # Create two x-values per week (left and right edges)
  tidyr::pivot_longer(
    cols = c(bar_left, bar_right),
    names_to = "edge",
    values_to = "x_outline"
  ) %>%
  dplyr::arrange(x_outline)


# -------------------------
# Build Plotly chart
# -------------------------
plot_ly() %>%
  
  # Bars: Registered deaths
  add_bars(
    data = df_observe_excess,
    x = ~week_num,
    y = ~DTHSREGPROV,
    name = "Registered Deaths",
    marker = list(color = nisra_blue),
    
    # Show actual date in tooltip
    hovertext = ~week_date,
    
    hovertemplate = paste(
      "<b>%{customdata}</b><br>",
      "Date: %{hovertext}<br>",
      "Registered Deaths: %{y}<extra></extra>"
    ),
    customdata = ~week_label
  ) %>%
  add_lines(
    data = df_expected_outline,
    x = ~x_outline,
    y = ~EXPDTHS,
    name = "Expected Deaths",
    line = list(color = nisra_navy, width = 3, shape = "hv"),
    hovertext = ~week_date,
    hovertemplate = paste(
      "<b>%{customdata}</b><br>",
      "Date: %{hovertext}<br>",
      "Expected Deaths: %{y}<extra></extra>"
    ),
    customdata = ~week_label
  ) %>%
  

  # -------------------------
  # Layout settings
  # ------------------------
  layout(
    margin = list(t = 70),
    
    # Transparent backgrounds (to match dashboard theme)
    plot_bgcolor = "rgba(0,0,0,0)",
    paper_bgcolor = "rgba(0,0,0,0)",
    
    # Custom x-axis (month labels)
    xaxis = list(
      title = "Month",
      tickmode = "array",
      tickvals = month_ticks$week_num,
      ticktext = month_ticks$x_tick_label,
      tickangle = 0,
      automargin = TRUE
    ),
    yaxis = list(
      title = "",
      rangemode = "tozero",
      ticks = "outside",
      ticklen = 8,
      tickcolor = '#FFFFFF',
      automargin = TRUE
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.25
    ),
    
    
    # Small spacing between bars
    bargap = 0.05,
    
    # Manual y-axis title placement (for styling control)
    annotations = list(
      list(
        text = "Number of deaths",
        x = -0.06,
        y = 1.05,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "bottom",
        font = list(size = 14)
      )
    )
  ) %>%
  
  # Remove Plotly mode bar (toolbar)

  config(displayModeBar = FALSE)
```

```{r download_reg_deaths, echo = FALSE}
# call make tables function to create chart download buttons
f_make_tables(
  data = df_observe_excess_dl,
  title = "RegisteredExpected",
  data_style = ns_comma,
  data_dir = here("outputs/figdata"),
)
```

:::
:::

<script src="scripts/info-boxes.js"></script>

<div id="reg-info-boxes"></div>

<script>
  document.addEventListener("DOMContentLoaded", () => {
    populateInfoBoxes(
  "reg-info-boxes",
  [
    "What is a registered death?",
    "What is an expected death?"
  ],
  [
    `
      <p>
        Registered deaths are counted based on the date the death is
        officially recorded with the General Register Office. There is a
        period of time between the date when a death occurs and the date it
        is registered. The number of days between death occurrence and
        registration varies, and can be greater in certain cases, for
        example when a death is referred to a coroner.
      </p>
    `,
    `
      <p>
        Expected deaths are an estimate of how many deaths would normally
        occur, based on current population and historical trends. These are
        used to compare with current figures. Expected deaths are calculated
        using statistical methods based on historical data trends. More
        information on the current method of estimating expected deaths and
        calculating excess deaths is available in the ONS Methodology Paper:
        <a
          href="https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/causesofdeath/datasets/estimatingexcessdeathsintheukmethodologychanges"
          target="_blank"
          rel="noopener noreferrer"
        >
          Estimating excess deaths in the UK, methodology changes - Office
          for National Statistics
        </a>.
      </p>
    `
  ]
);
  });
</script>

:::

</div>
</details>


<!-- --------------------------------------
     SECTION: Cause-specific deaths (Flu/Pneumonia & COVID-19)
     Displays weekly deaths where these conditions are mentioned
     on the death certificate
--------------------------------------- -->
<details open class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">

<!-- Section title -->
<h4 class="accordion-title">
Number of Flu/Pneumonia and COVID-19 related weekly deaths registered
</h4>

<span class="accordion-symbol" aria-hidden="true"></span>
</div>
</summary>

<div class="card-body">


<!-- Grid layout for responsive cards -->
::: {.grid}


<!-- --------------------------------------
     CARD 1: Flu / Pneumonia deaths
--------------------------------------- -->
::: {.g-col-12 .g-col-md-6}
::: {.card-padding .d-flex .h-100}
::: {.card .w-100}

::: {.card-header}
::: {style="text-align: left;"}
##### Flu/Pneumonia related weekly deaths registered
:::
:::

::: {.card-body .text-center}


<!-- Latest weekly value -->
<div class="fs-1 fw-bold" style="color:#00205b; text-align:left;">
`r flu_wdths`
</div>


<!-- Week-on-week change with direction arrow -->
<div class="fs-6 fw-bold" style="color:#000000; text-align:left;">
`r abs(flu_wdths_diff)` `r flu_wdths_arrow` `r flu_wdths_single_plural` than previous week
</div>


<!-- Previous week reference -->
<div class="fs-6 text-muted" style="text-align:left;">
Previous week: `r flu_wdths_prev`
</div>

:::
:::
:::
:::


<!-- --------------------------------------
     CARD 2: COVID-19 deaths
--------------------------------------- -->
::: {.g-col-12 .g-col-md-6}
::: {.card-padding .d-flex .h-100}
::: {.card .w-100}

::: {.card-header}
::: {style="text-align: left;"}
##### COVID-19 related deaths registered
:::
:::

::: {.card-body .text-center}

<div class="fs-1 fw-bold" style="color:#00205b; text-align:left;">
`r covid_wdths`
</div>

<div class="fs-6 fw-bold" style="color:#000000; text-align:left;">
`r abs(covid_wdths_diff)` `r covid_wdths_arrow` `r covid_wdths_single_plural` than previous week
</div>

<div class="fs-6 text-muted" style="text-align:left;">
Previous week: `r covid_wdths_prev`
</div>

:::
:::
:::
:::


<!-- --------------------------------------
     INFO BOX: Definitions
--------------------------------------- -->
::: {.g-col-12}
<div id="covid-info-boxes"></div>

<script>
  document.addEventListener("DOMContentLoaded", () => {
    populateInfoBoxes(
      "covid-info-boxes",
      [
        "What counts as a COVID-19 or flu/pneumonia related death?"
      ],
      [
        `
          <p>
            COVID-19 deaths include any mention of coronavirus, confirmed or
            suspected, anywhere on the death certificate. Flu or pneumonia
            related deaths include any mention of these conditions on the
            death certificate.
          </p>
        `
      ]
    );
  });
</script>
:::

:::

</div>
</details>



<!-- --------------------------------------
     SECTION: Sex breakdown of deaths 
     Displays weekly and year-to-date (YTD) deaths
     for males and females, including proportions
--------------------------------------- -->
<details closed class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">
<h4 class="accordion-title">
Sex breakdown of number of weekly deaths registered
</h4>

<span class="accordion-symbol" aria-hidden="true"></span>
</div>
</summary>

<div class="card-body">

::: {.grid}


<!-- --------------------------------------
     CARD 1: Female deaths
--------------------------------------- -->
::: {.g-col-12 .g-col-md-6}
::: {.card-padding .d-flex .h-100}
::: {.card .w-100}

::: {.card-header}
::: {style="text-align: left;"}
##### Number of female weekly deaths registered
:::
:::

::: {.card-body .text-center}


<!-- Weekly female deaths -->
<div class="fs-1 fw-bold" style="color:#00205b; text-align:left;">
`r female_wdths`
</div>


<!-- Year-to-date female deaths -->
<div class="fs-6 text-muted" style="text-align:left;">
Year to date total female deaths: `r female_ytd_wdths`
</div>

:::
:::
:::
:::

<!-- --------------------------------------
     CARD 2: Male deaths
--------------------------------------- -->
::: {.g-col-12 .g-col-md-6}
::: {.card-padding .d-flex .h-100}
::: {.card .w-100}

::: {.card-header}
::: {style="text-align: left;"}
##### Number of male weekly deaths registered
:::
:::

::: {.card-body .text-center}


<!-- Weekly male deaths -->
<div class="fs-1 fw-bold" style="color:#00205b; text-align:left;">
`r male_wdths`
</div>


<!-- Year-to-date male deaths -->
<div class="fs-6 text-muted" style="text-align:left;">
Year to date total male deaths: `r male_ytd_wdths`
</div>

:::
:::
:::
:::

:::


</div>
</details>


<!-- --------------------------------------
     SECTION: Age breakdown of weekly deaths
     Displays number of deaths by age band for most recent week
--------------------------------------- -->
<details closed class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">
<h4 class="accordion-title">
Age breakdown of number of weekly deaths registered
</h4>

<span class="accordion-symbol" aria-hidden="true"></span>
</div>
</summary>

<div class="card-body">

::: {.card-padding}
::: {.card}

<div class="rotate-screen">
  <p>Please rotate your screen for optimal experience</p>
</div>

::: {.card-body}

::: {#age-tabs .panel-tabset}

##### Weekly
```{r fig_age, fig-format: svg, echo=FALSE, out.width="100%", fig.alt="Bar chart showing deaths by age group for the latest week. The number of deaths increases with age, with the highest counts in the older age groups."}
# max x-axis range
max_x <- max(df_death_age$`Weekly Total`, na.rm = TRUE)

# Create horizontal bar chart
plot_ly(
  df_death_age,
  y = ~`Age band`,
  x = ~`Weekly Total`,
  type = "bar",
  orientation = "h",
  
  # Apply dashboard styling
  marker = list(
    color = nisra_blue,
    line = list(color = nisra_blue, width = 1)
  ),
  
  # Custom tooltip (cleaner than default hover)
  hoverinfo = "text",
  hovertemplate = "%{y}: %{x}<extra></extra>"
) %>%
  layout(
    margin = list(t = 70),  # gives space above plot for y-axis title
    plot_bgcolor = "rgba(0,0,0,0)",
    paper_bgcolor = "rgba(0,0,0,0)",
    xaxis = list(
      title = "Number of Deaths",
      range = c(0, max_x),
      autorange = FALSE,
      tick0 = 0,
      ntick = 10,
      showline = TRUE,
      zerolinecolor = "#666666"
    ),
    yaxis = list(
      autorange = "reversed",
      title = "",
      ticks = "outside",
      ticklen = 8,
      tickcolor = '#FFFFFF',
      automargin = TRUE
    ),
    
    # Custom label for y-axis ("Age")
    annotations = list(         # plot title above y-axis title
      list(
        text = "Age",
        x = -0.06,
        y = 1.05,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "bottom",
        font = list(size = 14)
      )
    )
  ) %>%
  
  # Remove Plotly toolbar for cleaner dashboard display
  config(displayModeBar = FALSE)

```

```{r download_age_weekly_deaths, echo = FALSE}
# call make tables function to create chart download buttons
f_make_tables(
  data = df_death_age_dl,
  title = "AgeWeekly",
  data_style = ns_comma,
  data_dir = here("outputs/figdata"),
)
```

##### Year to date
```{r fig_age_ytd, fig-format: svg, echo=FALSE, out.width="100%", fig.alt="Bar chart showing deaths by age group for the latest week. The number of deaths increases with age, with the highest counts in the older age groups."}
# max x-axis range
max_x <- max(df_age_ytd$`total_value`, na.rm = TRUE)

# Create horizontal bar chart
plot_ly(
  df_age_ytd,
  y = ~`Age band`,
  x = ~`total_value`,
  type = "bar",
  orientation = "h",
  marker = list(
    color = nisra_blue,
    line = list(color = nisra_blue, width = 1)
  ),
  
  # Custom tooltip for clarity
  hoverinfo = "text",
  hovertemplate = "%{y}: %{x}<extra></extra>"
) %>%
  layout(
    margin = list(t = 70),  # gives space above plot for y-axis title
    plot_bgcolor = "rgba(0,0,0,0)",
    paper_bgcolor = "rgba(0,0,0,0)",
    xaxis = list(
      title = "Number of Deaths",
      range = c(0, max_x),
      autorange = FALSE,
      tick0 = 0,
      ntick = 10,
      showline = TRUE,
      zerolinecolor = "#666666"
    ),
    yaxis = list(
      autorange = "reversed",
      title = "",
      ticks = "outside",
      ticklen = 8,
      tickcolor = '#FFFFFF',
      automargin = TRUE
    ),
    
    # Custom placement of y-axis title ("Age")
    annotations = list(         # plot title above y-axis title
      list(
        text = "Age",
        x = -0.06,
        y = 1.05,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "bottom",
        font = list(size = 14)
      )
    )
  ) %>%
  
  # Remove Plotly toolbar for cleaner presentation
  config(displayModeBar = FALSE)

```

```{r download_age_ytd_deaths, echo = FALSE}
# call make tables function to create chart download buttons
f_make_tables(
  data = df_age_ytd_dl,
  title = "AgeYTD",
  data_style = ns_comma,
  data_dir = here("outputs/figdata"),
)
```
:::

:::
:::



</div>
</details>


<!-- --------------------------------------
     SECTION: LGD (Local Government District) breakdown
     Displays weekly deaths geographically across NI districts
     using a map with ranking categories
--------------------------------------- -->

<details closed class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">
<h4 class="accordion-title">
Local government district (LGD) breakdown of number of weekly deaths registered
</h4>

<span class="accordion-symbol" aria-hidden="true"></span>
</div>
</summary>


<div class="card-body">

::: {.card-padding}
::: {.card}

<div class="rotate-screen">
  <p>Please rotate your screen for optimal experience</p>
</div>

::: {.card-body}

::: {.map-bg}

::: {#lgd-tabs .panel-tabset}

##### Weekly

```{r lgd deaths registered map, fig.alt="Map showing the number of deaths across Local Government Districts. Death counts vary by area, with some districts,like Belfast, showing higher numbers than others."}
#| out-width: 100%
#| fig-format: svg
#| fig-height: 4
#| dev.args: !expr list(bg = "#fff")

# Define color palette for ranking categories (dark to light)
category_colors <- c(
  "Highest Number of Deaths" = "#00205b",
  "2nd - 3rd Highest" = "#084594",
  "4th - 5th Highest" = "#2171b5",
  "6th - 7th Highest" = "#6baed6",
  "8th - 9th Highest" = "#bdd7e7",
  "10th - 11th Highest" = "#eff3ff"
)

# Plot with ranking-based legend
ggplot() +
  geom_sf(data = df_map, aes(fill = deaths_category), color = "grey40", size = 0.2) +
  scale_fill_manual(
    values = category_colors,
    labels = c(
      "Highest Number of Deaths" = "Highest",
      "2nd - 3rd Highest" = "",
      "4th - 5th Highest" = "",
      "6th - 7th Highest" = "",
      "8th - 9th Highest" = "",
      "10th - 11th Highest" = "Lowest"
    ),
    name = "",
    na.value = "grey90",
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      keyheight = 0.6,
      keywidth = 0.5
    )
  ) +
  geom_shadowtext(
    data = noncrowded_df %>% 
      filter(deaths_category %in% c("Highest", "2nd - 3rd Highest", "4th - 5th Highest")),
    aes(x = lng, y = lat, label = label_text),
    size = 2.5,
    fontface = "bold",
    show.legend = FALSE,
    lineheight = 0.8,
    color = "white",
    bg.color = "black"
  ) +
  geom_text(
    data = noncrowded_df %>% 
      filter(!deaths_category %in% c("Highest", "2nd - 3rd Highest", "4th - 5th Highest")),
    aes(x = lng, y = lat, label = label_text),
    size = 2.5,
    fontface = "bold",
    show.legend = FALSE,
    lineheight = 0.8,
    color = "black"
  ) +
  geom_point(
    data = crowded_df,
    aes(x = lng, y = lat),
    color = "black",
    size = 1.5
  ) +
  geom_text_repel(
    data = crowded_df,
    aes(x = lng, y = lat, label = label_text, color = label_colour),
    nudge_x = crowded_df$nudge_x,
    nudge_y = crowded_df$nudge_y,
    segment.color = "black",
    segment.size = 0.3,
    min.segment.length = 0,
    size = 2.5,
    fontface = "bold",
    direction = "both",
    box.padding = 0.4,
    max.overlaps = Inf,
    show.legend = FALSE,
    lineheight = 0.8
  ) +
  scale_color_identity() +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    plot.margin = margin(t = 0, b = 0),
    legend.position = c(1, 0.8),
    legend.margin = margin(l = 5),
    panel.background = element_rect(fill = "#fff", colour = NA),
    plot.background = element_rect(fill = "#fff", colour = NA),
  )
```

```{r download_lgd_weekly_deaths, echo = FALSE}
# call make tables function to create chart download buttons
f_make_tables(
  data = df_map_weekly_dl,
  title = "LGDWeekly",
  data_style = ns_comma,
  data_dir = here("outputs/figdata"),
)
```

##### Year to date

```{r lgd_deaths_ytd_registered_map, fig.alt="Map showing the number of deaths across Local Government Districts for the YTD. Death counts vary by area, with some districts,like Belfast, showing higher numbers than others."}
#| out-width: 100%
#| fig-format: svg
#| fig-height: 4
#| dev.args: !expr list(bg = "#fff")

# Define color palette for ranking categories (dark to light)
category_colors <- c(
  "Highest Number of Deaths" = "#00205b",
  "2nd - 3rd Highest" = "#084594",
  "4th - 5th Highest" = "#2171b5",
  "6th - 7th Highest" = "#6baed6",
  "8th - 9th Highest" = "#bdd7e7",
  "10th - 11th Highest" = "#eff3ff"
)

# Plot with ranking-based legend
ggplot() +
  geom_sf(data = df_ytd_map, aes(fill = deaths_category), color = "grey40", size = 0.2) +
  scale_fill_manual(
    values = category_colors,
    labels = c(
      "Highest Number of Deaths" = "Highest",
      "2nd - 3rd Highest" = "",
      "4th - 5th Highest" = "",
      "6th - 7th Highest" = "",
      "8th - 9th Highest" = "",
      "10th - 11th Highest" = "Lowest"
    ),
    name = "",
    na.value = "grey90",
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      keyheight = 0.6,
      keywidth = 0.5
    )
  ) +
  geom_shadowtext(
    data = noncrowded_ytd_df %>% 
      filter(deaths_category %in% c("Highest", "2nd - 3rd Highest", "4th - 5th Highest")),
    aes(x = lng, y = lat, label = label_text),
    size = 2.5,
    fontface = "bold",
    show.legend = FALSE,
    lineheight = 0.8,
    color = "white",
    bg.color = "black"
  ) +
  geom_text(
    data = noncrowded_ytd_df %>% 
      filter(!deaths_category %in% c("Highest", "2nd - 3rd Highest", "4th - 5th Highest")),
    aes(x = lng, y = lat, label = label_text),
    size = 2.5,
    fontface = "bold",
    show.legend = FALSE,
    lineheight = 0.8,
    color = "black"
  ) +
  geom_point(
    data = crowded_ytd_df,
    aes(x = lng, y = lat),
    color = "black",
    size = 1.5
  ) +
  geom_text_repel(
    data = crowded_ytd_df,
    aes(x = lng, y = lat, label = label_text, color = label_colour),
    nudge_x = crowded_ytd_df$nudge_x,
    nudge_y = crowded_ytd_df$nudge_y,
    segment.color = "black",
    segment.size = 0.3,
    min.segment.length = 0,
    size = 2.5,
    fontface = "bold",
    direction = "both",
    box.padding = 0.4,
    max.overlaps = Inf,
    show.legend = FALSE,
    lineheight = 0.8
  ) +
  scale_color_identity() +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    plot.margin = margin(t = 0, b = 0),
    legend.position = c(1, 0.8),
    legend.margin = margin(l = 5),
    panel.background = element_rect(fill = "#fff", colour = NA),
    plot.background = element_rect(fill = "#fff", colour = NA),
  )
```

```{r download_lgd_ytd_deaths, echo = FALSE}
# call make tables function to create chart download buttons
f_make_tables(
  data = df_map_ytd_dl,
  title = "LGDYTD",
  data_style = ns_comma,
  data_dir = here("outputs/figdata"),
)
```

:::

:::
:::

:::
<!-- --------------------------------------
     INFO BOX: Definitions
--------------------------------------- -->
::: {.g-col-12}
<div id="lgd-info-boxes"></div>

<script>
  document.addEventListener("DOMContentLoaded", () => {
    populateInfoBoxes(
      "lgd-info-boxes",
      [
        "How are deaths assigned to areas (LGDs)?"
      ],
      [
        `
          <p>
            Deaths are assigned based on the person’s usual residence 
            (which can include care homes). If the person did not usually live 
            in Northern Ireland, the death is assigned to the location of the 
            place of death.
          </p>
        `
      ]
    );
  });
</script>
:::

:::

</div>
</details>


<!-- --------------------------------------
     SECTION: Place of Death breakdown Displays number of deaths by place of occurrence
     for the most recent week
--------------------------------------- -->

<details closed class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">
<h4 class="accordion-title">
Place of Death breakdown of number of weekly deaths registered
</h4>

<span class="accordion-symbol" aria-hidden="true"></span>
</div>
</summary>

<div class="card-body" markdown="1">

::: {.card-padding}
::: {.card}

<!-- Prompt mobile users to rotate screen for better viewing -->

<div class="rotate-screen">
  <p>Please rotate your screen for optimal experience</p>
</div>

::: {.card-body}

::: {#pod-tabs .panel-tabset}

##### Weekly

```{r fig_place, fig-format: svg, echo=FALSE, out.width="100%", fig.alt="Bar chart showing deaths by place of occurrence for the latest week. Most deaths occur in hospitals and at home, with fewer deaths in care or nursing homes, and the lowest numbers in hospice and other locations."}
# max x-axis range
max_x <- max(df_pod$`Number of Deaths`, na.rm = TRUE)

# ensure order is descending
cat_order <- df_pod$`Place of death`[
  order(df_pod$`Number of Deaths`, decreasing = TRUE)
]


# -------------------------
# Create horizontal bar chart
# -------------------------
fig_place <- plot_ly(
  df_pod,
  y = ~`Place of death`,
  x = ~`Number of Deaths`,
  type = "bar",
  orientation = "h",  # horizontal bars
  
  # Apply consistent styling
  marker = list(
    color = nisra_blue,
    line = list(color = nisra_blue, width = 1)
  ),
  
  # Custom tooltip for clarity
  hoverinfo = "text",
  hovertemplate = "%{y}: %{x}<extra></extra>"
) %>%
  layout(
    
    # Space for custom axis title
    margin = list(t = 70),  # gives space above plot for y-axis title
    plot_bgcolor = "rgba(0,0,0,0)",
    paper_bgcolor = "rgba(0,0,0,0)",
    xaxis = list(
      title = "Number of Deaths",
      range = c(0, max_x),
      autorange = FALSE,
      tick0 = 0,
      ntick = 10,
      showline = TRUE,
      zerolinecolor = "#666666"
    ),
    yaxis = list(
      title = "",
      
      # Apply custom ordering (descending values)
      categoryorder = "array",
      categoryarray = cat_order,
      tickfont = list(size = 14),
      ticks = "outside",
      ticklen = 8,
      tickcolor = '#FFFFFF',
      automargin = TRUE
    ),
    annotations = list(         # plot title above y-axis title
      list(
        text = "Place of Death",
        x = -0.06,
        y = 1.05,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "bottom",
        font = list(size = 14)
      )
    )
  ) %>%
  
  # Remove Plotly toolbar for cleaner display
  config(displayModeBar = FALSE)


# Print plot
fig_place
```

```{r download_pod_weekly_deaths, echo = FALSE}
# call make tables function to create chart download buttons
f_make_tables(
  data = df_pod,
  title = "PlaceWeekly",
  data_style = ns_comma,
  data_dir = here("outputs/figdata"),
)
```

##### Year to date
```{r fig_place_ytd, fig-format: svg, echo=FALSE, out.width="100%", fig.alt="Bar chart showing deaths by place of occurrence for the year to date. Most deaths occur in hospitals and at home, with fewer deaths in care or nursing homes, and the lowest numbers in hospice and other locations."}
# max x-axis range
max_x <- max(df_ytd_pod$total_value, na.rm = TRUE)

# ensure order is descending
cat_order <- df_ytd_pod$`Place of death`[
  order(df_ytd_pod$total_value, decreasing = TRUE)
]

fig_place_ytd <- plot_ly(
  df_ytd_pod,
  y = ~`Place of death`,
  x = ~`total_value`,
  type = "bar",
  orientation = "h",
  marker = list(
    color = nisra_blue,
    line = list(color = nisra_blue, width = 1)
  ),
  hoverinfo = "text",
  hovertemplate = "%{y}: %{x}<extra></extra>"
) %>%
  layout(
    margin = list(t = 70),  # gives space above plot for y-axis title
    plot_bgcolor = "rgba(0,0,0,0)",
    paper_bgcolor = "rgba(0,0,0,0)",
    xaxis = list(
      title = "Number of Deaths",
      range = c(0, max_x),
      autorange = FALSE,
      tick0 = 0,
      ntick = 10,
      showline = TRUE,
      zerolinecolor = "#666666"
    ),
    yaxis = list(
      title = "",
      categoryorder = "array",
      categoryarray = cat_order,
      tickfont = list(size = 14),
      ticks = "outside",
      ticklen = 8,
      tickcolor = '#FFFFFF',
      automargin = TRUE
    ),
    annotations = list(         # plot title above y-axis title
      list(
        text = "Place of Death",
        x = -0.06,
        y = 1.05,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "bottom",
        font = list(size = 14)
      )
    )
  ) %>%
  config(displayModeBar = FALSE)

fig_place_ytd
```

```{r download_place_ytd_deaths, echo = FALSE}
# call make tables function to create chart download buttons
f_make_tables(
  data = df_ytd_pod_dl,
  title = "PlaceYTD",
  data_style = ns_comma,
  data_dir = here("outputs/figdata"),
)
```
:::
:::
:::

<!-- --------------------------------------
     INFO BOX: Definitions
--------------------------------------- -->
::: {.g-col-12}
<div id="pod-info-boxes"></div>

<script>
  document.addEventListener("DOMContentLoaded", () => {
    populateInfoBoxes(
      "pod-info-boxes",
      [
        "Place of Death note"
      ],
      [
        `
          <p>
            Care/Nursing homes - only includes deaths that occurred in those 
            settings. ‘Other’ includes deaths at non-usual residences or any 
            other locations not listed.
          </p>
        `
      ]
    );
  });
</script>

:::
:::

</div>

</details>


<!-- --------------------------------------
     SECTION: Registered deaths vs death occurrences (52-week view)
     Displays comparison between:
     - Registered deaths (based on registration date)
     - Death occurrences (based on actual date of death)
     over the most recent 52 weeks
--------------------------------------- -->

<details closed class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">
<h4 class="accordion-title">
Registered weekly deaths and death occurrences for the past 52 weeks
</h4>

<!-- Accordion expand/collapse icon -->
<span class="accordion-symbol" aria-hidden="true"></span>
</div>
</summary>

<div class="card-body">

::: {.card-padding}
::: {.card}

<!-- Prompt mobile users to rotate screen for better viewing -->

<div class="rotate-screen">
  <p>Please rotate your screen for optimal experience</p>
</div>

::: {.card-body}

```{r fig_overall_observe, fig-format: svg, echo=FALSE, out.width="100%", fig.alt="r chart of registered deaths by week with a line showing deaths that occurred in each week. The two series follow a similar pattern over time, with weekly values generally close but with some variation between registered and occurred deaths."}

# Create custom x-axis labels (months)
month_ticks <- df_reg_vs_occ %>%
  dplyr::group_by(month) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    # Identify where a new year begins
    year_marker = year != dplyr::lag(year, default = dplyr::first(year) - 1),
    # Display year below month when it changes
    x_tick_label = dplyr::if_else(
      year_marker,
      paste0(month_label, "<br>", year),
      month_label
    )
  )

# -------------------------
# Build Plotly chart
# -------------------------

plot_ly(df_reg_vs_occ, x = ~`Week Number`) %>%
  
  # Line 1: Registered deaths
  add_lines(
    y = ~`Registered Deaths`,
    name = "Registered Deaths",
    line = list(color = nisra_blue, width = 3),
    hovertext = ~week_ending_date,
    customdata = ~week_label_hover,
    hovertemplate = paste(
      "<b>Week %{customdata}</b><br>",
      "Date: %{hovertext}<br>",
      "Registered Deaths: %{y}<extra></extra>"
    )
  ) %>%
  
  # Line 2: Death occurrences (actual date of death)
  add_lines(
    y = ~`Deaths Occurred`,
    name = "Death occurrences",
    line = list(color = nisra_navy, width = 3),
    hovertext = ~week_ending_date,
    customdata = ~week_label_hover,
    hovertemplate = paste(
      "<b>Week %{customdata}</b><br>",
      "Date: %{hovertext}<br>",
      "Death occurrences: %{y}<extra></extra>"
    ) 
  ) %>%
  
  # -------------------------
  # Layout settings
  # -------------------------
  layout(
    margin = list(t = 70, b = 90),
    plot_bgcolor = "rgba(0,0,0,0)",
    paper_bgcolor = "rgba(0,0,0,0)",
    xaxis = list(
      title = "Month",
      tickmode = "array",
      tickvals = month_ticks$`Week Number`,
      ticktext = month_ticks$x_tick_label,
      tickangle = 0,
      automargin = TRUE
    ),
    yaxis = list(
      title = "",
      rangemode = "tozero",
      ticks = "outside",
      ticklen = 8,
      tickcolor = '#FFFFFF',
      automargin = TRUE
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.25
    ),
    annotations = list(
      list(
        text = "Number of Deaths",
        x = -0.06,
        y = 1.05,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "bottom",
        font = list(size = 14)
      )
    )
  ) %>%
  
  # Remove Plotly toolbar
  config(displayModeBar = FALSE)
```

```{r download_reg_vs_occ_deaths, echo = FALSE}
# call make tables function to create chart download buttons
f_make_tables(
  data = df_reg_vs_occ,
  title = "RegvsOcc",
  data_style = ns_comma,
  data_dir = here("outputs/figdata"),
)
```

:::
:::

<!-- --------------------------------------
     INFO BOX: Definitions
--------------------------------------- -->
::: {.g-col-12}
<div id="occ-info-boxes"></div>

<script>
  document.addEventListener("DOMContentLoaded", () => {
    populateInfoBoxes(
      "occ-info-boxes",
      [
        "Death registration verse death occurrence"
      ],
      [
        `
          <p>
            Occurred deaths (death occurrences) are based on the actual date the 
            person died, regardless of when the death was registered. This means 
            some most registered deaths may appear in a later week than when they 
            actually happened. Most deaths in Northern Ireland are registered 
            within 5 days. Delays can occur, especially around public holidays.
          </p>
        `
      ]
    );
  });
</script>

:::
:::

</div>
</details>


<!-- --------------------------------------
     SECTION: External links / related publications
     Provides users with links to detailed reports and related statistics across UK regions
--------------------------------------- -->
<details class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">
<h4 class="accordion-title">
Links to other publications on death registrations
</h4>

<span class="accordion-symbol" aria-hidden="true"></span>
</div>
</summary>

<div class="card-body">

- [Weekly death registrations in Northern Ireland](https://www.nisra.gov.uk/publications/weekly-death-registrations-northern-ireland-2026)

- Statistics on deaths registered in Northern Ireland, including weekly death registrations and excess deaths. [Death Statistics](https://www.nisra.gov.uk/statistics/deaths/death-statistics)

- Annual reports exploring specific causes of death, including alcohol, drug-related and suicide deaths in more detail. [Cause of Death statistics](https://www.nisra.gov.uk/statistics/births-deaths-and-marriages/cause-death)

- [Deaths registered weekly in England and Wales](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/datasets/weeklyprovisionalfiguresondeathsregisteredinenglandandwales)

- [Deaths registered weekly in Scotland](https://www.nrscotland.gov.uk/publications/deaths-registered-weekly-in-scotland/#)

</div>
</details>


<!-- --------------------------------------
     SECTION: Frequently Asked Questions (FAQs)
     Explains key concepts, definitions, and data caveats to help users interpret the statistics correctly
--------------------------------------- -->

<details class="card accordion">

<summary class="card-header accordion-header">
<div class="accordion-title-row">
<h4 class="accordion-title">
FAQs</h4>

<span class="accordion-symbol" aria-hidden="true"></span>
</div>
</summary>

<div class="card-body">

**1. What is the difference between registered, occurred and expected deaths?**  
**Registered deaths** are counted based on the date the death is officially recorded with the General Register Office.  
**Occurred deaths** (death occurrences) are based on the actual date the person died, regardless of when the death was registered.  
**Expected deaths** are an estimate of how many deaths would normally occur, based on historical trends. These are used to compare with current figures.


**2. Why might deaths appear in a different week than when they occurred?**  
Deaths are shown based on the registration date, not the date of death. This means some deaths may appear in a later week than when they actually happened. Most deaths in Northern Ireland are registered within 5 days. Delays can occur, especially around public holidays.


**3. How are expected deaths calculated?**  
Expected deaths are calculated using statistical methods based on historical data trends. More information on the current method of estimating expected deaths and calculating excess deaths is available in the ONS Methodology Paper at: [Estimating excess deaths in the UK, methodology changes - Office for National Statistics (ons.gov.uk)](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/causesofdeath/datasets/estimatingexcessdeathsintheukmethodologychanges)


**4. Are weekly figures final?**  
No — weekly figures are provisional. They may change as more data becomes available. Data are reviewed and updated quarterly as part of quality assurance.


**5. What counts as a COVID‑19 or flu-related death?**  
COVID‑19 deaths include any mention of coronavirus (confirmed or suspected) anywhere on the death certificate
Flu or pneumonia deaths include any mention of these conditions on the death certificate.


**6. How are deaths assigned to areas (LGDs)?**  
Deaths are assigned based on the person’s usual residence (which can include care homes). If the person did not usually live in Northern Ireland, the death is assigned to the place of death.


**7. What do the place of death categories mean?**  
Care/Nursing homes - only includes deaths that occurred in those settings. 'Other' includes deaths at non-usual residences or any other locations not listed.

</div>
</details>


<!-- --------------------------------------
     SCRIPT: Share functionality
     Enables users to share or copy link to dashboard
--------------------------------------- -->

<script>
function sharePage() {
  const shareData = {
    title: document.title,
    text: "Weekly Deaths Recorded in Northern Ireland",
    url: window.location.href
  };

  if (navigator.share) {
    navigator.share(shareData).catch(function(error) {
      console.log("Sharing cancelled or failed:", error);
    });
  } else if (navigator.clipboard) {
    navigator.clipboard.writeText(window.location.href).then(function() {
      alert("Page link copied to clipboard");
    });
  } else {
    prompt("Copy this link:", window.location.href);
  }
}
</script>
